Think again?\\ Optimal mental simulation tracks problem difficulty

### 2015-01-29 manuscript_ks.tex
KS edits to the manuscript

%% The question we're interested in: how many simulations do people take?
Imagine you are playing a video game---for example, Angry Birds or Desert Golf---in which you need to predict where an object will end up. How do you know where to aim, and how long do you spend thinking about each shot before you take it? Recent research in the domain of intuitive physics and mental simulation has revealed that people make predictions like these by running noisy physical simulations \cite{Smith:2013fc,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015}. However, while this research has investigated the \emph{mechanism} by which people make these types of predictions, there has been very little investigation into \emph{how} people actually use this mechanism. If people are running mental simulations, \emph{how many} simulations do they run?

%% What's already been done
According to previous research by \citeA{Vul:2014ba}, a sample-based agent should only take a small number of samples before making a judgment---perhaps even just one. While taking a single sample is clearly suboptimal for each individual judgment, over the long run, taking a small number of samples maximizes expected utility across a large number of judgments. Does this story also hold true for mental simulations? There is some evidence that points to ``yes'': \citeA{Battaglia2013} analyzed the variability of people's responses in tasks concerning towers of building blocks, and found that participants seemed to use between three and seven samples per judgment.

A second question concerns the actual strategy used to take samples. Do people always take exactly the same number of samples for each decision, or do they vary the number of samples based on problem difficulty? While \citeA{Vul:2014ba} looked at a variety of strategies, they concluded that regardless of strategy, the optimal solution was still roughly the same: the agent should take very few samples. However, that does not make the choice of strategy irrelevant. If the strategy tells the agent to vary the number of samples, then we should see variability in the agent's response times. Indeed, such variability can be observed in other types of decision-making tasks, and has been well-modeled by the \textit{sequential probability ratio test} \cite{wald1947sequential}, or SPRT \cite<e.g.>{Gold:2007fo,Ratcliff:2008ux,Bitzer:2014ea}. In SPRT, the agent takes samples until it reaches a threshold number of samples in support of one hypothesis over another hypothesis.

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=\textwidth]{figures/experiment.png}
        \caption{\textbf{Example experimental trial.} Each panel shows a different part of the trial. The left panel shows the initial screen presented to the participant. The middle panel shows the occluded ball, after observing the stimulus presentation. The faded gray line shows the path the ball took during the initial presentation. The right panel shows the final position of the ball, after observing the feedback. As in the middle panel, the faded gray line shows the path of the ball.}
        \label{fig:experiment}
    \end{center}
\end{figure*}

@@

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=\textwidth]{figures/hole_empirical_results.pdf}
        \caption{\textbf{Response times as a function of response.} Left subplot: each bar shows the proportion of participants saying that the ball will go in the hole for a particular feedback type ($x$-axis) and hole size (color). Middle subplot: like the left subplot, but the $y$-axis shows bootstrapped logarithmic means of response times. Right subplot: each point corresponds to a different stimulus, feedback type, and hole size.  The $x$-axis is the proportion of participants saying the ball will go in the hole, and the $y$ axis is the logarithmic mean response time.}
        \label{fig:pct-vs-rt}
    \end{center}
\end{figure*}

In order to determine whether people take multiples samples---and if they vary the number of samples---we designed an experiment in which people had to determine whether a ball would go into a hole (see Figure \ref{fig:experiment}). In particular, we designed the trials in this experiment such that some were harder than others.

@@

@@

The stimuli consisted of animations depicting a blue ball with a radius of 10px bouncing around in a box with dimensions 900px $\times$ 650px. All stimuli consisted of two separate animations. The first was the stimulus presentation, which had a duration of 0.775 seconds and depicted the ball moving in a particular direction. The second animation was the feedback, which picked up immediately where the first animation left off, and which had a duration of 1.5 seconds and depicted the ball either going into the hole or bouncing off the wall that contained the hole.\footnote{To enforce the constraint that the ball always travel the same distance during the feedback animation, the $x$-coordinate of the wall with the hole in it was allowed to vary across stimuli.} During the feedback animation, the ball could bounce on the other walls either 0, 1, or 2 times before going into the hole hitting the wall with the hole in it. In both animations, the ball had a velocity of 400px/s. Additionally, in the animations, the path that the ball had traveled so far was drawn with a faded gray line (see Figure \ref{fig:experiment}).

There were 48 different initial animations, and for each of these intial animations, there were four different types of feedback and two different hole sizes, giving a total of eight versions of each stimulus. The four feedback types were: ``far in'' (FI), where the ball went directly through the center of the hole; ``far miss'' (FM), where the ball missed the hole by a wide margin; ``close in'' (CI), where the ball just barely went through the hole; and `` close miss'' (CM), where the ball just barely missed the hole. The two hole sizes were 100px and 200px.

In order to ensure that participants never saw the same initial animation twice, we used a latin-square design of Initial Animation $\times$ Hole Type $\times$ Hole Size. Thus, each participant saw each initial animation exactly once, each hole type exactly 12 times, and each hole size exactly 24 times.

@@

The experiment was divided into two phases: the training phase, and the experimental phase. During the training phase, participants made judgments on the seven instruction trials in order to get used to the task. During the experimental phase, participants made judgements on the 48 experimental trials, presented in a random order, as well as the eight control trials, which were also shown in a random order, but interspersed with the experimental trials such that every 8th trial was a control trial.

On each trial, participants were shown the scene, including the initial position of the ball and the location of the hole. Participants were instructed to press the ``space'' key to begin the trial. Immediately upon pressing ``space'', the initial stimulus presentation began. As soon as the initial stimulus animation concluded, a gray box was drawn over the screen, occluding the ball (but not the line depicting the path it had traveled so far; this was left in as a reminder to participants of where the ball had come from). Participants were asked, ``will the ball go in the hole?'', and were instructed to press `q' to respond in the affirmative, and `p' to respond in the negative. Immediately after responding, text appeared saying ``Correct!'' or ``Incorrect.'', depending on the participants' response. Additionally, the gray occluder was removed, and participants were shown the feedback animation. After the feedback animation was complete, the final frame of the animation remained on the screen until participants pressed ``space'' to advance to the next trial.

@@

For all analyses of response time, we computed averages using bootstrapped logarithmic means (exponential of the mean of the log response times), using 1000 bootstrap samples. Participants had an average response time of \AvgRT{} milliseconds (\StdRT{}ms), excluding catch trials. There were significant effects of both feedback type (\RTHoleClass{}) and hole size (\RTHoleSize{}) on response time, as well as an interaction between feedback type and hole size (\RTFull{}). However, the hole size was only significant in the case of the ``far in'' (FI) trials ($p<0.001$), in which the ball went straight through the center of the hole. Figure \ref{fig:pct-vs-rt} (middle subplot) shows the response times as a function of feedback type and hole size.

@@

As shown in Figure \ref{fig:pct-vs-rt} (right subplot), we found a clear effect of certainty on people's response times. In particular, participants were on average faster to respond on trials that they were more certain about, and slower to respond to trials that they were unsure about. This tradeoff mirrors the same tradeoff found in SPRT: decisions for which $p\approx0.5$ are slower, because it takes longer to get to one threshold versus the other. Thus, it is clear that people are not spending a fixed amount of time thinking about each trial; rather, they spend more time thinking about the ``hard'' trials than the ``easy'' trials.

@@

\subsection{The ``Noisy Newton'' hypothesis}

There is a growing body of evidence that people reason about physical scenes by running noisy simulations. This hypothesis, referred to as the ``Noisy Newton'' hypothesis, states that people have approximate knowledge of physical laws instantiated in a runnable model of intuitive physics. Using this model, they can extrapolate the future of physical scenes by running a series of noisy simulations \cite{Smith:2013fc,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015}. In particular, \citeA{Smith:2013fc} investigated the various sources of uncertainty in these simulations, finding that people's judgments were best captured by a model that took into account both perceptual uncertainty and dynamics uncertainty.

>>>>

%% The question we're interested in: how many simulations do people take?
Imagine you are playing the game Angry Birds, where the goal is to launch a bird to knock down a tower. As you aim the bird, you need to predict where the bird will go to decide whether your shot will have the desired effect. How do you know where to aim, and how long do you spend thinking about each shot before you take it? If you shoot the bird without spending time to consider where it will go, your shot will likely be far off. But on the other hand, if you spend too long aiming, over time your predictions will only get marginally better, and you eat into the time you have to play until your next meeting. Making this physical prediction requires a tradeoff that underlies many cognitive tasks: spending time and cognitive resources to increase confidence about what will transpire next, versus using that information to make decisions in the moment. In this paper, we study whether people efficiently trade off between information seeking and exploitation in a physical prediction task.

Recent research in the domain of physical reasoning has revealed that people make predictions like these by running noisy physical simulations \cite{Smith:2013fc,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015} - the noisy Newton theory \cite{Sanborn2013}. However, while this research has investigated the \emph{mechanism} by which people make these types of predictions, there has been very little investigation into \emph{how} people actually use this mechanism. If we are running mental simulations, how to we choose how many sample simulations to run?

%% What's already been done
According to previous research by \citeA{Vul:2014ba}, a sample-based agent should only take a small number of samples before making a judgment so long as there is any cost to taking samples---perhaps even just one. While taking a single sample clearly provides a worse chance of making a good decision than taking multiple samples, over the long run, taking a small number of samples maximizes expected utility across a large number of judgments. Does this story also hold true for mental simulations? There is some evidence that points to ``yes'': \citeA{Battaglia2013} analyzed the variability of people's responses in tasks concerning towers of building blocks, and found that participants seemed to use between three and seven samples per judgment.

A second question concerns the actual strategy used to take samples. Do people always take exactly the same number of samples for each decision, or do they vary the number of samples based on problem difficulty? While \citeA{Vul:2014ba} looked at a variety of strategies, they concluded that regardless of strategy, the optimal solution was still roughly the same: the agent should take very few samples. However, that does not make the choice of strategy irrelevant---certain strategies require different numbers of samples depending on how variable those samples are. For instance, using the \textit{sequential probability ratio test}, or SPRT\cite{wald1947sequential}, the agent takes samples until it reaches a threshold number of samples in support of one hypothesis over another. SPRT has been used to explain the variability in a number of decision-making tasks \cite<e.g.>{Gold:2007fo,Ratcliff:2008ux,Bitzer:2014ea}, and so is a candidate mechanism for taking physical simulation samples. But if this strategy tells the agent to vary the number of samples, then we should see variability in the agent's response times as more or fewer samples are taken.

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=\textwidth]{figures/experiment.png}
        \caption{\textbf{Example experimental trial.} Each panel shows a different part of the trial. \emph{Left:} the initial screen presented to the participant. \emph{Middle:}  the screen is occluded after observing the stimulus presentation. The faded gray line shows the path the ball took during the initial presentation. \emph{Right:} the final position of the ball, after observing the feedback. As in the middle panel, the faded gray line shows the path of the ball.}
        \label{fig:experiment}
    \end{center}
\end{figure*}

@@

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=\textwidth]{figures/hole_empirical_results.pdf}
        \caption{\textbf{Response times as a function of response.} \emph{Left:} each bar shows the proportion of participants saying that the ball will go in the hole for a particular feedback type ($x$-axis) and hole size (color). \emph{Middle:} like the left subplot, but the $y$-axis shows bootstrapped logarithmic means of response times. \emph{Right:} each point corresponds to a different stimulus, feedback type, and hole size.  The $x$-axis is the proportion of participants saying the ball will go in the hole, and the $y$ axis is the logarithmic mean response time.}
        \label{fig:pct-vs-rt}
    \end{center}
\end{figure*}

In order to determine whether people take physical simulation samples in a way consistent with SPRT, we designed an experiment in which people made a binary judgment about what will happen in the future---will a ball traveling through a computer screen go through a hole or not (see Figure \ref{fig:experiment})? In particular, we designed the trials such that some were harder than others. If people are using an SPRT sampling strategy, we would expect people to take longer (due to more sampling) when it was more ambiguous whether the ball would go through the hole.

@@

On each trial, participants were shown the scene, including the initial position of the ball and the location of the hole. Participants were instructed to press the ``space'' key to begin the trial. Immediately upon pressing ``space'', the initial stimulus presentation began. As soon as the initial stimulus animation concluded, a gray box was drawn over the screen, occluding the ball (but not the line depicting the path it had traveled so far; this was left in as a reminder to participants of where the ball had come from). Participants were asked, ``will the ball go in the hole?'', and were instructed to press `q' to respond in the affirmative, and `p' to respond in the negative. Immediately after responding, text appeared saying ``Correct!'' or ``Incorrect.'', depending on the participants' response. Additionally, the gray occluder was removed, and participants were shown feedback animation of the path of the ball. After the feedback animation was complete, the final frame of the animation remained on the screen until participants pressed ``space'' to advance to the next trial.

During the experiment, participants made judgements on 48 experimental trials, presented in a random order, as well as eight control trials, which were also shown in a random order, but interspersed with the experimental trials such that every 8th trial was a control trial. In addition, participants were given seven instruction trials prior to the experiment to familiarize them with the procedure.

@@

The stimuli consisted of animations depicting a blue ball with a radius of 10px bouncing around in a box with dimensions 900px $\times$ 650px. All stimuli consisted of two separate animations. The first was the stimulus presentation, which had a duration of 0.775 seconds and depicted the ball moving in a particular direction. The second animation was the feedback, which picked up immediately where the first animation left off, and which had a duration of 1.5 seconds and depicted the ball either going into the hole or bouncing off the wall that contained the hole.\footnote{To enforce the constraint that the ball always travel the same distance during the feedback animation, the $x$-coordinate of the wall with the hole in it was allowed to vary across stimuli.} During the feedback animation, the ball could bounce on the other walls either 0, 1, or 2 times before going into the hole or hitting the wall with the hole in it. In both animations, the ball had a velocity of 400px/s. Additionally, in the animations, the path that the ball had traveled so far was drawn with a faded gray line (see Figure \ref{fig:experiment}).

There were 48 different initial animations, equally balanced by number of bounces during feedback (16 each for 0, 1, and 2 bounces). For each of these iniital animations, there were four different types of feedback and two different hole sizes, giving a total of eight versions of each stimulus. The four feedback types were: ``far in'' (FI), where the ball went directly through the center of the hole; ``far miss'' (FM), where the ball missed the hole by a wide margin; ``close in'' (CI), where the ball just barely went through the hole; and `` close miss'' (CM), where the ball just barely missed the hole. The two hole sizes were 100px and 200px.

In order to ensure that participants never saw the same initial animation twice, we used a latin-square design of Initial Animation $\times$ Hole Type $\times$ Hole Size. Thus, each participant saw each initial animation exactly once, each hole type exactly 12 times, and each hole size exactly 24 times. This also ensured that the ball would go through the hole exactly half the time, so that participants would not be biased to respond either way based on statistical contingencies.

@@

@@

For all analyses of response time, we computed averages using bootstrapped logarithmic means (exponential of the mean of the log response times), using 1000 bootstrap samples. Participants had an average response time of \AvgRT{} milliseconds (\StdRT{}ms), excluding catch trials. There were significant effects of both feedback type (\RTHoleClass{}) and hole size (\RTHoleSize{}) on response time, as well as an interaction between feedback type and hole size (\RTFull{}). However, there was only evidence for a difference in reaction times by hole size in the case of the ``far in'' (FI) trials ($p<0.001$), in which the ball went straight through the center of the hole. Figure \ref{fig:pct-vs-rt} (middle subplot) shows the response times as a function of feedback type and hole size.

@@

As shown in Figure \ref{fig:pct-vs-rt} (right subplot), we found a clear effect of certainty on people's response times. In particular, on trials when there was more agreement amongst participants (the fraction saying 'in' is closer to $0$ or $1$), participants were faster to respond than when responses were more variable. This tradeoff mirrors the same tradeoff found in SPRT: decisions for which $p\approx0.5$ are slower, because it on average takes more samples to get to one threshold versus the other. Thus, it is clear that people are not spending a fixed amount of time thinking about each trial; rather, they spend more time thinking about the ``hard'' trials than the ``easy'' trials.

In addition, we found that response time is affected by the number of bounces during feedback; participants were slower to respond as more bounces were added. While this may be partially modulated by difficulty---people make more variable predictions as bounces are added \cite{Smith:2013fc}, and more variable trials have longer reaction times---there appears to be an additional time cost. \TODO{Can we add statistics here? E.g., how much additional time, and something showing that this is above and beyond difficulty (intercept differences with quadratic fits)?} This suggests that it takes additional time to resolve a collision between the ball and a wall in the course of a simulation. We will use this theory to further explain expected reaction times in \TODO{Add section here}.

@@

\subsection{The ``noisy Newton'' hypothesis}

There is a growing body of evidence that people reason about physical scenes by running noisy simulations. This hypothesis, referred to as the ``noisy Newton'' hypothesis, states that people have approximate knowledge of physical laws instantiated in a runnable model of intuitive physics. Using this model, they can extrapolate the future of physical scenes by running a series of noisy simulations \cite{Smith:2013fc,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015}. In particular, \citeA{Smith:2013fc} investigated the various sources of uncertainty in these simulations, finding that people's judgments were best captured by a model that took into account both perceptual uncertainty (noise in exactly where objects are and their motion trajectories) and dynamic uncertainty (stochastic noise added to account for unknowable variation in motion---e.g., a textured floor would cause a ball rolling along it to deviate from a straight line).
